Parsing Postfix log files

### 2007-11-25 doc/logparser.tex
Finished Carl's first set of corrections; some of them are added as XXX items, but at least I've gotten the vast majority done.

@@

\begin{abstract}

@@

Section~\ref{database schema} describes the database schema used by the
parser, explaining in detail the tables used for storing the data gleaned
from the logs and the table that stores the rules.

Section~\ref{rules} discusses the parsing rules in detail, describing the
table in the database used to hold the rules and explaining their
structure, referring to an example rule and sample data it matches
successfully against.  This section concludes with a discussion of rule
efficiency concerns.

@@

Section~\ref{parsing coverage} analyses the coverage the parser achieves
over a set of 93 log files taken from a mail server handling mail for over
1000 users.  Coverage is described both in terms of the fraction of lines
parsed and the fraction of mails and connections successfully reconstructed
by the parsing algorithm.

@@

    \item \SQL{} is reasonably standard and many people will already be
        familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn.  Although every
        vendor implements a different dialect of \SQL{}, the basics are the
        same everywhere (analogous to the overall similarities and minor
        differences between British English, American English and
        Australian English).  A good introduction to \SQL{} can be found
        at~\cite{sql-for-web-nerds}, others
        are~\cite{w3schools-sql-tutorial, sqlcourse.com}.

@@

The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity many virus and/or spam sending programs fail to implement it
properly, so requiring strict adherence to the protocol specification is
beneficial in protecting against spam and viruses.\footnote{This violates
the principle of \textit{Be liberal in what you accept, and conservative in
what you send\/} from \RFC{}~760~\cite{rfc760}, but unfortunately that
principle was written in a friendlier time. XXX CARL WANTS ME TO SUGGEST A
NEW PRINCIPAL.}  A typical \SMTP{} conversation resembles the following
(the lines starting with a three digit number are sent by the server, all
other lines are sent by the client):

@@

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive and extensible optional
anti-spam restrictions, allowing an administrator to deploy those
restrictions which they judge suitable for their site's needs, rather than
a fixed set chosen by Postfix's author.  These restrictions can be
selectively applied, combined and bypassed on a per-client, per-recipient
or per-sender basis, allowing varying levels of severity and/or
permissiveness.  Postfix leverages simple lookup tables to support
arbitrarily complicated user-defined sequences of restrictions and
exceptions, with policy servers\footnote{Policy servers will be explained
in section~\refwithpage{policy servers}.} as the ultimate in flexibility.
Administrators can also supply their own rejection messages to make it
clear to senders why exactly their mail was rejected.  Unfortunately this
flexibility has a cost: complexity in the logs generated.  While it is easy
to use standard Unix text processing utilities to determine the fate of an
individual email, following the journey an email takes through Postfix can
be quite difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of
the time the journey is simple, but the other 10\% of the time requires
90\% of the code.\footnote{These numbers don't have a solid scientific
basis, they're based on gut feeling from writing and debugging the
software.}

@@

In some ways this task is similar to reverse engineering or replicating a
black box program based solely on its inputs and outputs.  Although the
source code is available, reading and understanding it would require a
significant investment of time:

\begin{tabular}[]{llll}

    Postfix 2.2.11  & Postfix 2.3.8   & Postfix 2.4.0 &                   \\
    71548           & 82224           & 83965         & lines of code     \\
    60962           & 67146           & 68675         & lines of comments \\
    16117           & 17647           & 18069         & lines are blank   \\
    148627          & 167017          & 170709        & lines in total    \\

@@

@@

>>>>

% This may keep hyperref happy about the Abstract entry in the table of
% contents.
\newcounter{dummy}

@@

\refstepcounter{dummy}
\addcontentsline{toc}{section}{Abstract}
\begin{abstract}

@@

This algorithm requires a database for storing both the rules used when
parsing and the results gleaned from parsing.  The database schema used is
described in section~\ref{database schema}, explaining in detail the tables
used for storing the data gleaned from the logs and the table that stores
the rules.

Section~\ref{rules} discusses the parsing rules in detail, explaining the
purpose and usage of each field in a rule, referring to an example rule and
sample data it matches successfully against.  This section concludes with a
discussion of rule efficiency concerns.

@@

Section~\ref{parsing coverage} analyses the coverage the parser achieves
over a set of 93 log files taken from a mail server handling mail for over
1000 users.  Coverage is described both in terms of the fraction of lines
parsed and the fraction of mails and connections successfully reconstructed
by the parsing algorithm, including dealing with false negatives and a
discussion of the difficulties in identifying false positives.

@@

    \item \SQL{} is reasonably standard and many people will already be
        familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn.  Although every
        vendor implements a different dialect of \SQL{}, the basics are the
        same everywhere (analogous to the overall similarities and minor
        differences between British English, American English and
        Australian English).  A good introduction to \SQL{} can be found
        at~\cite{sql-for-web-nerds}, others
        are~\cite{w3schools-sql-tutorial, sqlcourse.com}.  Depending on the
        database in use there may be tools available which reduce or remove
        the requirement to know \SQL{}; for SQLite (the default database
        used by the implementation) there are several
        available~\cite{sqlite-guis}.

@@

The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity many virus and/or spam sending programs fail to implement it
properly, so requiring strict adherence to the protocol specification is
beneficial in protecting against spam and viruses.\footnote{This violates
the principle of \textit{Be liberal in what you accept, and conservative in
what you send\/} from \RFC{}~760~\cite{rfc760}, but unfortunately that
principle was written in a friendlier time. XXX CARL WANTS ME TO SUGGEST A
NEW PRINCIPAL AND POSSIBLY AN APPENDIX OF SPAM RELATED ISSUES\@.}  A
typical \SMTP{} conversation resembles the following (the lines starting
with a three digit number are sent by the server, all other lines are sent
by the client):

@@

An example deviation from the protocol:

\begin{verbatim}
220 smtp.example.com ESMTP
HELO client.example.com
250 smtp.example.com
MAIL FROM: Alice N. Other alice@example.com
501 5.1.7 Bad sender address syntax
RCPT TO: Bob in Sales/Marketing bob@example.com
503 5.5.1 Error: need MAIL command
DATA
503 5.5.1 Error: need RCPT command
Message headers and body sent here.
.
502 5.5.2 Error: command not recognized
QUIT
221 2.0.0 Bye
\end{verbatim}

This client is so poorly written that not only does it present the sender
and recipient addresses improperly, it ignores the error messages returned
by the server and carries on regardless.

@@

Postfix is a \MTA{} with the following design aims (in order of
importance): security, flexibility of configuration, scalability, and high
performance.  It features extensive, extensible, optional anti-spam
restrictions, allowing an administrator to deploy those restrictions which
they judge suitable for their site's needs, rather than a fixed set chosen
by Postfix's author.  These restrictions can be selectively applied,
combined and bypassed on a per-client, per-recipient or per-sender basis,
allowing varying levels of severity and/or permissiveness.  Postfix
leverages simple lookup tables to support arbitrarily complicated
user-defined sequences of restrictions and exceptions, with policy
servers\footnote{Policy servers will be explained in
section~\refwithpage{policy servers}.} as the ultimate in flexibility.
Administrators can also supply their own rejection messages to make it
clear to senders why exactly their mail was rejected.  Unfortunately this
flexibility has a cost: complexity in the logs generated.  While it is easy
to use standard Unix text processing utilities to determine the fate of an
individual email, following the journey an email takes through Postfix can
be quite difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of
the time the journey is simple, but the other 10\% of the time requires
90\% of the code.\footnote{These numbers don't have a solid scientific
basis, they're based on gut feeling from writing and debugging the
software.}

@@

In some ways this task is similar to reverse engineering or replicating a
black box program based solely on its inputs and outputs.  Although the
source code is available,\footnote{Reading and understanding the source
code would require a significant investment of time; the source code for
Postfix 2.4.6 is 16MB.} there are advantages to treating Postfix as a black
box while developing the parser:

    \item The parser is developed using real world logs rather than the
        idealised logs someone would naturally envisage reading the source
        code.

    \item The source code cannot accurately communicate the variety of
        orderings in which log lines are written to the log file, as
        process scheduling interferes with it.

    \item The parser acts as a second source of information, with the
        information gathered from empirical evidence.  An interesting
        project would be to compare the empirical knowledge inherent in the
        parsing algorithm with the documentation and source code of
        Postfix.

@@

XXX REFER TO THE RULES/PARSER SPLIT ELSEWHERE RATHER THAN EXPLAINING IT
AGAIN HERE\@.

XXX``CATEGORISE THE COMPLICATIONS SECTIONS IN WORDS HERE'' --- CARL\@.

XXX DISCUSS THE OUTPUT, OR LACK THEREOF, IN GREATER DETAIL, AFTER RERUNNING
\LMA{}\@.

@@

XXX ADD SOMETHING ABOUT FALSE POSITIVES AND NEGATIVES\@.
